\chapter{Publications}
This chapter presents a list of the author's journal, conference proceedings, and other publications and dissemination events. Some of them are directly related to the chapters of the PhD manuscript, whereas others belong to other projects where the author has participated. 

\section*{Included in the thesis}
\subsection*{Published journal papers}

\begin{itemize}
	%\baselineskip=12pt
	
	\item[\textbf{J1}] \'{I}. Munn\'{e}-Collado, F. M. Apr\`{a}, P. Olivella-Rosell, R. Villafafila-Robles, A. Sumper, ``The potential role of flexibility during peak hours on greenhouse gas emissions: a life cycle assessment of five targeted national electricity grid mixes," \textit{Energies}, vol.~12, no.~23, Nov. 2019. doi: 10.3390/en12234443
	
\end{itemize}

\subsection*{Submitted journal papers}
\begin{itemize}
	\item [\textbf{J2}] \'{I}. Munn\'{e}-Collado, P. Pinson, M. Arag\"{u}\'{e}s-Pe\~{n}alba, A. Sumper, "Aggregated flexibility forecast using online recursive maximum likelihood kernel density estimation" April 2021, Submitted to:  Status:
	\item [\textbf{J3}] OPF MV networks 
\end{itemize}

\subsection*{Conference papers}

\begin{itemize}
	\item[\textbf{C1}] \'{I}. Munn\'{e}-Collado, P. Lloret-Gallego, P. Olivella-Rosell, R. Villafafila-Robles, S. \O{}. Ottesen, R. Gallart-Fernandez, V. Palma-Costa, A. Sumper, ``System architecture for managing congestions in distributions grids using flexibility," 25th International Conference on Electricity Distribution, June 2019.
\end{itemize}

\subsection*{Book chapters}

\begin{itemize}
	\item[\textbf{BC1}] \'{I}. Munn\'{e}-Collado, E. Bullich-Massagu\'{e}, M. Arag\"{u}\'{e}s-Pe\~{n}alba, P. Olivella-Rosell ``Local and Micro Power Markets," in A. Sumper (ed) Micro and Local Power Markets, John Wiley \& Sons, pp.~37-97, 2019. \\doi: 10.1002/9781119434573.ch2
\end{itemize}

\subsection*{Data bases}

\begin{itemize}
	\item [\textbf{DB1}] \'{I}. Munn\'{e}-Collado, P. Pinson, "Aggregated flexibility forecast using online recursive maximum likelihood kernel density estimation - Associated data and code", Mendeley Data, March 2021,  \\doi: 10.17632/t92mmtm4gs.1
\end{itemize}

\section*{Not included in the thesis}

\subsection*{Published journal papers}

\begin{itemize}
	
		\item [\textbf{J4}] P. Olivella-Rosell, P. Lloret-Gallego, \'{I}. Munn\'{e}-Collado, R. Villaf\'{a}fila-Robles, A. Sumper, S. Ottesen, J. Rajasekharan, B. Bremdal, ``Local flexibility market design for aggregators providing multiple flexibility services at distribution network Level," \textit{Energies}, vol.~11, no.~4, p.~822, Apr.~2018. doi: 10.3390/en11040822
	\item [\textbf{J5}]  S. Barja-Mart\'{i}nez, F. Rucker, M. Arag\"{u}\'{e}s-Pe\~{n}alba, R. Villaf\'{a}fila-Robles, \'{I}. Munn\'{e}-Collado, P. Lloret-Gallego, "A novel hybrid home energy management system considering electricity cost and greenhouse gas emissions minimization" \textit{IEEE Transactions on Industry Applications}, February~2021
	\item [\textbf{J6}] H. Kazmi, \'{I}. Munn\'{e}-Collado, F. Mehmood, T. A. Syed, J. Driesen, "Towards data-driven energy communities: a review of open-source datasets, models and tools". \textit{Renewable and Sustainable Energy Reviews}, June 2021.
\end{itemize}

\subsection*{Submitted journal papers}
\begin{itemize}
	\item [\textbf{J7}] S. Barja-Mart\'{i}nez, M. Arag\"{u}\'{e}s-Pe\~{n}alba,\'{I}. Munn\'{e}-Collado, P. Lloret-Gallego, E. Bullich-Massagu\'{e}, R. Villaf\'{a}fila-Robles "Artificial intelligence techniques for enabling big data services in distribution networks: a review". Submitted to Renewable and Sustainable Energy Reviews. Status: Under Review
	\item [\textbf{J8}] H. Kazmi, \'{I}. Munn\'{e}-Collado, K. Tidriri, L. Norrdstrom, F. Gielen, J. Driesen, "Data Science and Energy: Some lessons from Europe on higher education course design and delivery", Submitted to Harvard Data Science Review, July 2020. Status: Under Review
\end{itemize}

	
\subsection*{Book chapters}

\begin{itemize}
	\item[\textbf{BC2}] \'{I}. Munn\'{e}-Collado, P. Olivella-Rosell, A. Sumper, ``Power Market Fundamentals," in A. Sumper (ed) Micro and Local Power Markets, John Wiley \& Sons, pp.~1-35, 2019. doi: 10.1002/9781119434573.ch1
\end{itemize}	


\subsection*{Conference papers}

\begin{itemize}
	\item [\textbf{C2}] M.Myllysilta, I. Deviatkin, J. Sampa,  \'{I}. Munn\'{e}-Collado, S. Tuurala, A. Hentunen, T. Pajula  "Life Cycle Assessment of First- and Second-Life Lithim-Ion Batteries: Implications from Existing Studies," Going Green CARE INNOVATION 2018 - Schoenbrunn Palace Conference Center, Vienna, Austria, November 2018
	\item [\textbf{C3}] S. Barja-Mart\'{i}nez, \'{I}. Munn\'{e}-Collado, P. Lloret-Gallego, M. Arag\"{u}\'{e}s-Pe\~{n}alba, R. Villaf\'{a}fila-Robles "A Novel Home Energy Management System Environmental-based with LCA Minimization", 2020 IEEE International Conference on Environment and Electrical Engineering (EEIC), June 2020
	\item [\textbf{C4}] \'{I}. Munn\'{e}-Collado, A. Bov\'{e} Salat, D. Montesinos-Miracle, R. Villaf\'{a}fila-Robles "A data acquisition pipeline for home energy management systems" III Congreso Iberoamericano de Ciudades Inteligentes, November 2020

\end{itemize}

\subsection*{Local journals}
\begin{itemize}
	\item [\textbf{C5}] O. Boix, \'{I}. Munn\'{e}-Collado, "Tecnologia vestible: combinant disseny, art i tecnologia", Revista de ecnologia, 8 (2020) p. 23-32,  \\doi: 10.2436/20.2004.01.25
\end{itemize}

\subsection*{Education}

\begin{itemize}
	\item [\textbf{E1}] \'{I}. Munn\'{e}-Collado, Cristian Chill\'{o}n-Ant\'{o}n, Jos\'{e} Ignacio Bustamante-Vargas, Marc Jen\'{e}-Vinuesa, Andreas Sumper "Control and Automation for the Efficient Use of Energy" Learning Material, September 2018, September 2019.	
	\item [\textbf{E2}] O. Boix, C. Lamp\'{o}n, E. Sanz, \'{I}. Munn\'{e}-Collado, "Tecnolog\'{i}a Vestible a tu alcance. Wearable Technology. MOOC Course", February 2021	.
\end{itemize}


\subsection*{Presentations in other events}
\begin{itemize}
	\item [\textbf{P-C1}] Presentation of "Local Smart Grids Architecture and Life Cycle Assessment" in EMPOWER General Assembly, Wolpertshausen, September 2017.		
	\item [\textbf{P-C2}] Presentation of "Interactive learning in higher education" in InnoEnergy Teachers Conference 2018 , Santa Cruz de Tenerife, April 2018.	
	\item [\textbf{P-C3}] Presentation of "Local Markets and Flexibility Services" in InnoEnergy MSc SENSE and Smart Cities Workshop , Barcelona, Spain, October 2018.	
	\item [\textbf{P-C4}] Presentation of "Local Markets and Flexibility Services: EMPOWER and INVADE" in Smart Grids Workshop, Barcelona, Spain, December 2018.	
	\item [\textbf{P-C5}] Presentation of "The environmental impact of smart grids: Life Cycle Assessment" in Master in Sustainability, Escola Enginyeria Bes\`{o}s Est (EEBE), Barcelona, Spain, December 2019, May 2020, December 2021, May 2021.
	\item [\textbf{P-C6}] Presentation of "Data Science in the energy field: Workshop" in Ironhack Full-time Data Analytics Bootcamp, Barcelona, Spain, December 2019.
	\item [\textbf{P-C7}] Presentation of "Home and Small buildings automation: our role as end-users in the energy transition" in InnoEnergy Community Speakers Series, Barcelona, Spain, April 2019.
	\item [\textbf{P-C8}] Presentation of "Innovation in higher education" in 1st InnoEnergy Local Teachers Conference, Barcelona, Spain, March 2019.
	\item [\textbf{P-C9}] Presentation of "Learning Analytics and Flipped Classroom approach at UPC" in MIT-Station 1 Workshop , Boston, USA, May 2019.	
	\item [\textbf{P-C10}] Presentation of "Introduction to Data Science in the energy field" in InnoEnergy MSc SENSE Workshop, Barcelona, Spain, October 2019.
	\item [\textbf{P-C11}] Presentation of "Life Cycle Assessment of Smart Grids and electricity markets" in INVADE Final Event, Barcelona, Spain, December 2019.
	\item [\textbf{P-C12}] Presentation of "Aggregated flexibility forecast: The service of BD4OPEM Project" in ELMA Seminars, Kongens Lyngby, Denmark, April 2021.
	\item [\textbf{P-C13}] Presentation of "Learning Analytics: The case study of Control and automation at UPC" in 2nd InnoEnergy Local Teachers Conference, Barcelona, Spain, February 2020.
	\item [\textbf{P-C14}] Presentation of "Introduction to Data Science in the energy field" in InnoEnergy MSc SENSE Workshop, Barcelona, Spain, December 2020.
	\item [\textbf{P-C15}] Presentation of "Flexibility in distribution networks" in 1st Seminar on Dones At\`{o}miques, Barcelona, Spain, May 2021.
\end{itemize}

\subsection*{Supervised bachelor and master thesis}

\begin{itemize}
	\item [\textbf{T1}] M. Guari, ``Dimensionat d'un gasificador de biomassa de petita pot\`{e}nia i estudi d'implantaci\'{o} en una microxarxa mitjan \c{c}ant Matpower", June 2017.
	\item [\textbf{T2}] F. Apr\`{a}, ``Environomical analysis of peak hours electricity production in targeted European countries", June 2019.
	\item [\textbf{T3}] K.Beehuspoteea,  ``Impact factors of heat generation units for zoned temperature controlled in office buildings", June 2019.
	\item [\textbf{T4}] A. Quattrone, ``Development of flexibility device models for a microgrid laboratory test", June 2019. 
	\item [\textbf{T5}] N. Condorelli, ``Evaluation and forecast of CO2 emissions in the
electricity sector for European targeted countries" March 2020.
	\item [\textbf{T6}] P. Plana, ``Analysis of measures to increment the share of renewable energy in distribution grids" April 2020.
	\item [\textbf{T7}] A.Vega, ``Proyecto de generaci\'{o}n fotovoltaica distribuida y almacenamiento energ\'{e}tico compartido con bater\'{i}as Li-Ion de segunda vida" May 2020.
	\item [\textbf{T8}] A.E. Marerro Le\'{o}n, ``Micro-red aplicable a pol\'{i}gonos industriales mediante bater\'{i}as Li-Ion de segunda vida" May 2020.
	\item [\textbf{T9}] A. Bov\'{e} Salat, ``Optimal scheduling of flexible assets under a HEMS for prosumers' economic savings" June 2020. 
	\item [\textbf{T10}] M. Ferran, ``Power flow tool for active distribution grids and flexibility analysis", May 2021. 
	\item [\textbf{T11}] X. Vel\'{a}zquez, ``An\'{a}lisis del impacto de las energ\'{i}as renovables en los sobrecostes de generaci\'{o}n del sistema el\'{e}ctrico insular de Tenerife mediante la aplicaci\'{o}n de ML", May 2021. 
	 
\end{itemize}

\subsection*{Published technical reports}

\begin{itemize}
	\item[\textbf{TR1}] E. F. B\o{}dal, P. Crespo-del-Granado, H. Farahmand, M. Korp\aa{}s, P. Olivella-Rosell, I. Munn\'{e}-Collado, P. Lloret-Gallego, ``INVADE Deliverable 5.1 Challenges in distribution grid with high penetration of renewables," June 2017. doi: 10.5281/zenodo.853271
	\item[\textbf{TR2}] I. Munn\'{e}-Collado, ``EMPOWER Deliverable 3.5 Life-Cycle Analysis", December 2017.
	\item[\textbf{TR3}] I. Munn\'{e}-Collado, F. Apr\`{a}, ``INVADE Deliverable 3.4 Draft Life-Cycle Asessment", December 2018.
	\item[\textbf{TR4}] P. Olivella-Rosell, P. Lloret-Gallego, L. Haupt, S. Barja-Mart\'{i}nez, I. Munn\'{e}-Collado, S. Bjarghov, et al., ``INVADE Deliverable 5.4 Advanced Optimal Battery operation and control algorithm", December 2018.
	\item[\textbf{TR5}] I. Munn\'{e}-Collado, F. Apr\`{a}, M. Myllysilta, ``INVADE Deliverable 3.5 INVADE Pilot-sites Life-Cycle Assessment", December 2019.	
	\item[\textbf{TR6}] S. Barja-Mart\'{i}nez, I. Munn\'{e}-Collado, M. Jen\'{e}-Vinuesa, P. Lloret-Gallego et al., ``BD4OPEM Deliverable 4.1 BD4OPEM technologies and services. First Version", June 2021.	
\end{itemize}
