\renewcommand{\abstractname}{Thesis development framework}
\begin{abstract}

The research presented in this thesis has been involved in the following R\&D projects and entities:
\begin{itemize}
\item EMPOWER "Local electricity retail markets for prosumer smart grid power services project", European Union's Horizon 2020 Research and Innovation programme under Grant Agreement No. 646476. 
\item INVADE "Smart system of renewable energy storage based on integrated EVs and batteries to empower mobile, distributed and centralised energy storage in the distribution grid", European Union's Horizon 2020 Research and Innovation programme under Grant Agreement No. 731148. 
\item BD4OPEM "Big Data for Open Innovation Energy Marketplace", European Union's Horizon 2020 Research and Innovation programme under Grant Agreement No. 872525.
\item e-Alma "Sistema de Almacenamiento de energ\'{i}a h\'{i}brido para plantas de generaci\'{o}n el\'{e}ctrica renovables", Ministerio de Economia, Industria y Competitividad (Spanish government) under Grant Agreement No.~ENE2017-692 86493-R.
\item FLEXRED "Flexibilidad de  los  recursos  energ\'{e}ticos  distribuidos  parar  optimizar la operaci\'{o}n de las redes de distribuci\'{o}n", Ministerio de Ciencia, Innovaci\'{o}n y Universidades (Spanish government) under Grant Agreement No. RTI2018-099540. 
\item Centre d'Innovaci\'{o} Tecnol\`{o}gica en Convertidors Est\`{a}tics i Accionaments, CITCEA-UPC, Spain. 
\item Universitat Polit\`{e}cnica de Catalunya, UPC, Spain.
\item Danish Technical University, DTU, Denmark. 
\end{itemize}

\vspace*{2.5mm}

The thesis has been financially supported by the following entities and grants:  

\begin{itemize}
\item Ag\`{e}ncia de Gesti\'{o} d'Ajuts Universitaris i de Recerca, FI-AGAUR 2017.
\item The European Institute of Innovation \& Technology (EIT) InnoEnergy Ph.D. School 2017.
\item The European Institute of Innovation \& Technology (EIT) InnoEnergy Master Program and Teachers Excellence Awards 2019.
\end{itemize}

\end{abstract}